Large squeezing in Bose-Einstein condensates with large numbers of atoms

### 2013-07-09 large_number_squeezing_with_atoms.tex
Results conclusions viable draft

\title{Large squeezing in multicomponent trapped Bose-Einstein condensates}

@@

\begin{abstract}
We examine the feasibility of creating and measuring large relative number squeezing in multicomponent trapped Bose-Einstein condensates.  In the absence of multimode effects, this squeezing can be arbitrarily large, but a range of processes limit the measurable squeezing in realistic trap configurations.  We examine these processes, and suggest methods to mitigate them. We conclude that high levels of squeezing with large numbers of atoms is feasible, but can realistically only be achieved in traps with even density profiles in at least two dimensions. We also introduce a method of maximising the measurable squeezing by using a pi-pulse during the process to improve spatial mode-matching.
\end{abstract}

@@

\section{Introduction}
\label{sectionIntroduction}
\begin{itemize}
  \item General non-classical state generation with atoms
  \item Talk about number squeezing with atoms
  \item Link to interferometry
  \item Motivate why large number squeezing
  \item Point out no one has done it before, only low number squeeze
  \item Lay out what sections we have and what we've done
\end{itemize}

Quantum optics led to creating nonclassical states after the laser was invented \cite{???}. The experimental realization of the BEC \cite{???} led to the ability to start recreating these types of states with massive particles.

One of the advantages of atoms over photons is that they interact with each other, unlike photons, which allows nonlinear interaction effects without all the fluffery that optics requires, like crystals etc.

It has been recognized that the intrinsic nonlinearities of atoms could lead to effects like quadrature squeezing \cite{kheruntsyanET2005, johnssonET2007, haineET2009, doeringET2010}. Can also get number squeezing using spin squeezing techniques \cite{kitagawaET1993, gross2010, sinatra2012}, molecular dissocation of BECs \cite{kheruntsyanET2002, davisET2008, savage2007, ogren2010}, four-wave mixing \cite{perrin2008, ogren2009, jascula2010}, or double-well potentials \cite{esteve2008, maussang2010} or mixing quadrature-squeezed states using Ramsey interferometry \cite{haineET2009}.

{\bf Note that the FWM stuff is using the s-wave scatter sphere from condensate collisions, so pretty useless for interferometry}

The creation of these non-classical atomic states has relevence for fundamental tests of quantum mechanics such as the EPR paradox \cite{pu2000, sorensen2001, haine2005,zhao2007, ogren2010}, as well as atom inteferometry \cite{sorensenET1999, kuzmichET2000, meyerET2001, liebfriedET2004, roosET2006, doeringET2010, gross2010}

While these techniques have resulted in impressive levels of experimentally detectable number squeezing \cite{gross2010}, they have been limited to small numbers of atoms, typically a few hundred.

To really win in the interferometry game, you need good squeezing for large numbers of atoms. The reason people have used small numbers of atoms is that it allows you to move into the single-mode regime, where the trap is tight enough such that all higher modes are frozen out \cite{LiET2009}

If this is not the case multimode stuff needs to be considered, and that's what we do in this paper.

Most previous simulations of this sort of stuff tend to be in 1D or 2D for simplicity \cite{kheruntsyanET2002, kheruntsyanET2005, johnssonET2007, haineET2009}. They reduce the problem from 3D to 1D by integrating out one or two dimensions \cite{???}. Semi-classically this isn't a problem, because you can assume the modal dynamics have symmetry. However as we show in this paper, when dealing with QFT problems this naive approach can give you the wrong answer. We can do full 3D simulations because we're cool and we have XMDS2 \cite{dennis2013}.

\section{Model and two mode solutions}
\label{secTwoModeAnalytic}
Our goal is to describe the number squeezing possibilities of a BEC comprising two relevant internal states denoted by $|a\rangle$ and $|b\rangle$.  The Hamiltonian for a pair of coupled atomic internal modes is the spatial integral of the Hamiltonian density, which is given by
\begin{equation}
\hat{\mathcal{H}} = \sum_{j} \hat{\psi}_j^{\dagger}\left(-\frac{\hbar^2}{2 m}\nabla^2+V_j\right)\hat{\psi}_j 
          + \sum_{i j}\frac{U_{i j}}{2} \hat{\psi}_i^{\dagger} \hat{\psi}_j^{\dagger} \hat{\psi}_j \hat{\psi}_i
          + \kappa \hat{\psi}_a^{\dagger} \hat{\psi}_b + \kappa^* \hat{\psi}_b^{\dagger}  \hat{\psi}_a
\label{eqFieldHamiltonian}
\end{equation}
where $\hat{\psi}_j(\mathbf{x})$ is the annihilation field operator for a particle at position $\mathbf{x}$ and in internal state $|j\rangle$, $m$ is the mass of the atoms, $V_j(\mathbf{x})$ is the trapping potential for atoms in internal state $|j\rangle$, $U_{ij}$ describe the various inter- and intra-state nonlinearities, and $\kappa$ describes the coupling between the two states.

We begin by considering a simplified version of the problem, assuming that only one of these spatial modes is relevant for each of the condensate's components.  This is done by writing the multimode atomic field operators in terms of a set of basis functions $\hat{\psi}_a({\mathbf{r}},t) = \sum_n u_{n}({\mathbf{r}}) \hat{a}_{n}(t)$ and $\hat{\psi}_b({\mathbf{r}},t) = \sum_n u_{n}({\mathbf{r}}) \hat{b}_{n}(t)$, where $\hat{a}_{n}$ and $\hat{b}_{n}$ annihilate a particle with normalised spatial mode $u_n$ and internal state $|a\rangle$ or $|b\rangle$ respectively. With this expansion, the nonlinear term in the multimode Hamiltonian for the self-interaction of the $|a\rangle$ internal state becomes
\begin{equation}
\hat{H} = \frac{U_{aa}}{2} \int dV \sum_{nmpq} u_n^* u_m^* u_p u_q \, \hat{a}^{\dagger}_{n} \hat{a}^{\dagger}_{m} \hat{a}_{p} \hat{a}_{q} 
\end{equation}
When the system remains predominantly in one mode, we can approximate this sum as the single term $\hat{H} = \hbar\chi_{aa} \hat{a}^{\dagger} \hat{a}^{\dagger} \hat{a} \hat{a}/2$ where
\begin{equation}
\chi_{aa} = \frac{1}{\hbar} U_{aa} \int |u_{i}({\mathbf{r}})|^4 \, dV.
\label{eqChiUequivalence}
\end{equation}
We proceed similarly for the other terms.  In the parameter regimes we are interested in, the detuning between the two spatial modes will be irrelevant, so setting our zero of energy appropriately, we are left with only a standard Kerr-type nonlinearity arising from atom-atom interactions, as well as a linear coupling between the two modes.  The Hamiltonian governing this reduced system is given by
\begin{equation}
\hat{H} = \hbar\frac{\chi_{aa}}{2} \hat{a}^{\dagger} \hat{a}^{\dagger} \hat{a} \hat{a}
          + \hbar\chi_{ab} \hat{a}^{\dagger} \hat{a} \hat{b}^{\dagger} \hat{b}
          + \hbar\frac{\chi_{bb}}{2} \hat{b}^{\dagger} \hat{b}^{\dagger} \hat{b} \hat{b}
          + \hbar\Omega (\hat{a}^{\dagger} \hat{b} + \hat{b}^{\dagger}  \hat{a} )
\label{eqTwoModeHamiltonian}
\end{equation}
where the $\chi_{ij}$ describe the various inter- and intra-mode nonlinearities, and $\Omega$ couples the two modes and allows for population transfer between them.

The system is prepared with all the population in $|a\rangle$, and vacuum in $|b\rangle$. We choose state $|a\rangle$ to be occupied as a coherent state with average number $\langle \hat{a}^{\dagger} \hat{a} \rangle = N$, where $N$ is the total particle number in our system.  As our simulation and experiment are insensitive to the initial phase of the coherent state, this is equivalent to a mixture of coherent states of uncertain phase, which is also equivalent to mixture of Poissonian-distributed number states.  This state is consistent with BEC coherence experiments \cite{Hadzibabic2004}.

At time $t=0$, the coupling $\Omega$ is switched on and applied until time $t=t_1$, at which point it is turned off, resulting in a portion of the population being transferred into mode $|b\rangle$. We assume this transfer process fast relative to the nonlinear energy time scale, ensuring that, as the initial states were coherent states, after the population transfer both $|a\rangle$ and $|b\rangle$ are still described by coherent states with mean number $\langle \hat{a}^{\dagger} \hat{a} \rangle = n_a$ and $\langle \hat{b}^{\dagger} \hat{b} \rangle = n_b$ respectively.

The coupling $\Omega$ is then switched off until time $t_2$. During this period the atoms interact solely through the nonlinear terms, with no population being transferred. Due to the form of the Hamiltonian, the number variance remains unchanged, but the quadrature variances in the two modes change over time, which can lead to quadrature squeezing \cite{johnssonET2007}.

After this hold time $\tau_{\mathrm{hold}} = t_2 - t_1 $, the coupling $\Omega$ is switched back on until time $t_3$, with a phase shift $\phi$ compared to its first application. During this last stage, the two modes exchange population and the quadrature variance is converted into number variance, in the same way that homodyne measurements are used in quantum optics to convert quadrature squeezing into number squeezing, which can be directly measured. In this interpretation, $\phi$ is the relative phase of the strong local oscillator, which allows specific phase angles of quadrature squeezing to be examined.  This experiment can also be interpreted as a Ramsey interferometer with a final beam splitter phase of $\phi$.

The entire sequence of pulses and the resulting populations are shown schematically in Figure~\ref{figPulseScheme}.

@@

We might speculate that the effects of these first two problems would be lessened by trap geometries that lead to near-constant density profiles for the BEC.  This should enable higher squeezing.  We cannot make strong conclusions without also considering the effects of coupling between spatial modes, however, which leads to the third potential problem: Coupling between modes is inevitable in the presence of a nonlinearity, meaning that the modes cannot be analyzed independently. Coupling between modes will disturb mode-matching, as well as mixing fields with different phase evolution, which can rapidly destroy squeezing.

@@

These stochastic equations take the form
\begin{eqnarray}
\frac{d \phi_a}{dt} = -\frac{i}{\hbar}\left(-\frac{\hbar^2}{2 m}\nabla^2+V_a(\mathbf{x}) + U_{a a} \left( \left|\phi_{a}\right|^2 -\frac{1}{dV}\right) + U_{a b} \left( \left|\phi_b\right|^2 -\frac{1}{2dV}\right)  \right)\phi_a   + \kappa \phi_b  \nonumber \\
\frac{d \phi_b}{dt} = -\frac{i}{\hbar}\left(-\frac{\hbar^2}{2 m}\nabla^2+V_b(\mathbf{x}) + U_{b b} \left( \left|\phi_{b}\right|^2 - \frac{1}{dV} \right) + U_{a b} \left( \left|\phi_a\right|^2 -\frac{1}{2dV} \right)   \right)\phi_b   + \kappa^* \phi_a
  \label{eqStochasticEquations}
\end{eqnarray}
where $dV$ is the volume element of the grid on which the simulation is carried out, and the terms proportional to $1/dV$ correspond to a vacuum corrections. It is a peculiarity of the Wigner representation that, up to the vacuum correction terms, Eqs.~(\ref{eqStochasticEquations}) look identical to those of the coupled semi-classical nonlinear Schr{\"{o}}dinger equation. The quantum statistical nature of the Wigner equations enters due to the fact that Eqs.~(\ref{eqStochasticEquations}) are run many times and stochastically averaged, with each run using different set of random initial conditions describing (in this case) the noise on a coherent state.

@@

\section{Effects of trap geometries on squeezing in 1D}

@@

\section{Using a $\pi$-pulse to improve mode-matching}
Modal mismatch is generated by the different atomic potentials seen by the two internal states.  A straightforward method of minimising this effect is to use a $\pi$-pulse to swap the populations of the internal states after half of the evolution. As can be seen by Eqs~(\ref{eqa1}) and (\ref{eqb1}), such a pulse also leaves the quantum statistics of each state unchanged. In the limit where each state remains in a single mode, and exactly half of the atoms are in each state, this technique ensures that both modes have identical mode shapes after the hold time. In practice this will never work perfectly --- due to the density distributions of $\psi_a(\mathbf{r})$ and $\psi_a(\mathbf{r})$ not reflecting the ground state, resulting in non-identical environments for the two modes, some modal dynamics will occur. Nonetheless, in many regimes, such a $\pi$-pulse can produce some improvement. More importantly, as this scheme accepts that the two modes will change shape, and attempts to make them change in a similar fashion, it allows the possibility of much longer hold times. That is, if we are no longer constrained to hold times short enough such that the mode shapes of the atoms in states $|a \rangle$ and $|b \rangle$ do not change significantly, much more squeezing can be extracted from the system.

These points are illustrated in Fig.~\ref{piPulseFig}, which shows the best squeezing obtainable from a specific system with and without the $\pi$-pulse, as well as the extra squeezing that can be obtained by allowing a longer hold time as well as the $\pi$-pulse.
\begin{figure}
    \centering
    \includegraphics[width=8cm]{figures/pi_pulse.eps}
    \caption{Effect of a $\pi$-pulse halfway through the hold time. Common parameters used were $n_a = n_b =2 \times 10^5$, $\chi_{aa}=0.03\,\text{s}^{-1}$, $\chi_{ab}=\chi_{bb}=0$. Solid red line: Analytic two-mode solution, $\tau_{\mathrm{hold}}=3\times 10^{-4}$s, $\phi=3.0$ (10.9dB squeezing); Solid blue line: no $\pi$-pulse, $\tau_{\mathrm{hold}}=3\times 10^{-4}$s, $\phi=3.0$ (4.9dB squeezing);  Dashed blue line: with $\pi$-pulse, $\tau_{\mathrm{hold}}=3\times 10^{-4}$s, $\phi=3.05$ (5.8dB squeezing); Dash-dotted blue line: with $\pi$-pulse and the longer hold time $\tau_{\mathrm{hold}}=5\times 10^{-4}$s allowed by the application of the pulse, $\phi=3.87$ (7.2dB squeezing).}
    \label{piPulseFig}
\end{figure}

@@

To minimise the damaging effects of the non-condensed fraction it is better to operate in a regime where the total occupation of the non-zero momentum modes $N_{\mathbf{k}\neq 0}$ is small, which is also fortuitously the regime in which the Bogoliubov theory is a good approximation.  To better understand when $N_{\mathbf{k}\neq 0}/N$ will be small, we approximate the sum in \eqref{eqCondensateDepletion} with an integral.  In 1D Eq.~\eqref{eqCondensateDepletion} is
\begin{align}
  \frac{N_{\mathbf{k}\neq 0}(t)}{N} &= \frac{(n_a \chi_{aa} \tau_\text{hold} \cos\theta)^2}{N} \sum_{k \neq 0} \sinc^2\left(\sqrt{\omega^0_{k}(\omega^0_{k}+2 \chi_{aa}n_a)}\tau_\text{hold}\right).
\end{align}
We wish to approximate this sum by the integral
\begin{align}
  \frac{N_{\mathbf{k}\neq 0}(t)}{N} &\approx \frac{(n_a \chi_{aa} \tau_\text{hold} \cos\theta)^2}{N} \int_{-\infty}^{+\infty} \sinc^2 \left(\sqrt{\omega^0_{k}(\omega^0_{k}+2 \chi_{aa}n_a)}\tau_\text{hold}\right) \frac{d k}{\Delta k}, \label{eqCondensateDepletionFirstIntegralApproximation}
\end{align}
where $\Delta k = 2\pi/L$ and $L$ is the length of the 1D box condensate.  This approximation will be valid in the limit that $\omega_{k_\text{min}} \tau_\text{hold} \ll 1$, where $k_\text{min} = 2\pi/L$ as for small $k$ the allowed values $k = n_x 2\pi /L$ will closely sample the region $\xi = \sqrt{\omega_k^0(\omega_k^0 + 2 \chi_{aa} n_a)} \tau_\text{hold} < \pi$ where $\sinc^2(\xi)$ is significant.

>>>>

\title{Large squeezing in Bose-Einstein condensates with large numbers of atoms}

@@

\begin{abstract}
We examine the feasibility of creating and measuring large relative number squeezing in multicomponent trapped Bose-Einstein condensates.  In the absence of multimode effects, this squeezing can be arbitrarily large, but a range of processes limit the measurable squeezing in realistic trap configurations.  We examine these processes, and suggest methods to mitigate them. We conclude that high levels of squeezing with large numbers of atoms is feasible, but can realistically only be achieved in traps with particular trap geometries.  We also introduce a method of maximising the measurable squeezing by using a pi-pulse during the process to improve spatial mode-matching.
\end{abstract}

@@

\section{Introduction}
\label{sectionIntroduction}

Quantum degenerate ultracold atomic sources have been used to study fundamental processes such as quantum phase transitions \cite{Leggett2001,Buluta2009}, quantum non-equilibrium thermodynamics \cite{Kinoshita:2006,Barnett:2011} and entanglement of massive particles \cite{Kheruntsyan:2005}.  Their ability to be highly isolated from their environment and controlled precisely by optical, magnetic and rf fields makes them an ideal platform for engineering and manipulating non-trivial quantum states.   In quantum optics, the most common way to produce entangled states has been through squeezing, and it has been recognised that the intrinsic nonlinearities of atoms naturally lead to quadrature squeezing \cite{kheruntsyanET2005, johnssonET2007, haineET2009, doeringET2010}.  Number squeezing can be achieved with spin squeezing techniques \cite{kitagawaET1993, gross2010, sinatra2012}, molecular dissociation of Bose-Einstein Condensates (BECs) \cite{kheruntsyanET2002, davisET2008, savage2007, ogren2010}, four-wave mixing \cite{perrin2008, ogren2009, jascula2010}, double-well potentials \cite{esteve2008, maussang2010}, or mixing quadrature-squeezed states using Ramsey interferometry \cite{haineET2009}.
The creation of these non-classical atomic states has relevance for fundamental tests of quantum mechanics such as the EPR paradox \cite{pu2000, sorensen2001, haine2005,zhao2007, ogren2010}, as well as atom inteferometry \cite{sorensenET1999, kuzmichET2000, meyerET2001, liebfriedET2004, roosET2006, doeringET2010, gross2010}.
While these techniques have resulted in significant experimentally detectable number squeezing \cite{gross2010}, they have been limited to small numbers of atoms, typically a few hundred.

Ultracold atomic interferometers are also reaching state of the art sensitivities in precision measurements \cite{Robins2012}. While all current atom interferometry is limited by technical noise sources, it appears that BEC-based coherent sources offer the best chance of overcoming some of these issues, for the same reasons that optical lasers are typically preferred for optical interferometry \cite{Szigeti2012}.  If these technical noise sources can be controlled, atom interferometers will be limited by atomic shot noise, which leads to a sensitivity that scales as the square root of the atomic flux.   From that point, sensitivity will only be improved by increasing the atomic flux, and/or using non-classical quantum states such as squeezed states to go below the shot noise limit.  Producing squeezed states for metrological purposes is therefore only relevant when it can be achieved in a context of large numbers of atoms.  This paper examines the generation of atomic squeezing using the inherent Kerr-like nonlinearities of ultracold atoms, and demonstrates a trap geometry in which metrologically relevant levels of squeezing are present in trap geometries that are compatible with large atomic number.

The creation, detection and application of squeezing requires precise mode-matching, and is therefore very sensitive to uncontrolled coupling between spatial modes of the atomic field.  Atomic squeezing experiments have typically attempted to operate in the single-mode regime, where the trap is tight enough such that all higher modes are frozen out \cite{LiET2009}.  However, tight traps limit the total number of atoms, as the 3-body recombination rate places an effective maximum atomic density in the gas, and well before that limit, the interatomic interactions break the single-mode operation.  Both of these effects limit the number of atoms that can be effectively squeezed, so we will examine a broader range of trap geometries, and calculate the dynamics in the presence of multiple spatial modes.
Many similar previous simulations have been performed in one or two spatial dimensions, by integrating out some of the remaining dimensions \cite{kheruntsyanET2002, kheruntsyanET2005, johnssonET2007, haineET2009}.  For simulations of semiclassical fields this can be a reasonable approximation, but it can ignore spontaneous scattering processes in full quantum field theory calculations, so we take care to include full 3D simulations.

In section \ref{secTwoModeAnalytic} we examine the ideal behaviour when the system is reduced to a single spatial mode, extracting a full analytic solution to the squeezing for arbitrary \textbf{whatever it is we extended, along with reference of what was covered previously.}  In section \ref{sec:MMdescription} we explore the idea of how multimode behaviour will become important in physical systems, and introduce our model for simulating higher dimensional trap geometries.  Sec.~\ref{sec:1Dsim} analyses one-dimensional models, highlighting some of the issues regarding optimising squeezing parameters and mode-matching in squeezing experiments.  Sec.~ref{sec:pipulse} shows that the application of a $\pi$-pulse during the process can dramatically improve mode-matching issues.  Sec.~\ref{sec:2D3Dsqueezing} investigates measurable squeezing in 2D and 3D, and uses a Bogoliubov analysis to explain the  scaling with dimensionality.  Sec.~\ref{sectionConclusion} describes the conclusion that large volume, flat-bottomed, three-dimensional traps, which are obviously compatible with large atom number, can produce extremely high degrees of squeezing.

\section{Model and two mode solutions}
\label{secTwoModeAnalytic}
Our goal is to describe the number squeezing possibilities of a BEC comprising two relevant internal states denoted by $|a\rangle$ and $|b\rangle$.  The Hamiltonian for a pair of coupled atomic internal modes is the spatial integral of the Hamiltonian density:
\begin{equation}
\hat{\mathcal{H}} = \sum_{j\in \{a,b\}} \hat{\psi}_j^{\dagger}\left(-\frac{\hbar^2}{2 m}\nabla^2+V_j\right)\hat{\psi}_j 
          + \sum_{i,j}\frac{U_{i j}}{2} \hat{\psi}_i^{\dagger} \hat{\psi}_j^{\dagger} \hat{\psi}_j \hat{\psi}_i
          + \kappa \hat{\psi}_a^{\dagger} \hat{\psi}_b + \kappa^* \hat{\psi}_b^{\dagger}  \hat{\psi}_a,
\label{eqFieldHamiltonian}
\end{equation}
where $\hat{\psi}_j(\mathbf{x})$ is the annihilation field operator for a particle at position $\mathbf{x}$ and in internal state $|j\rangle$, $m$ is the mass of the atoms, $V_j(\mathbf{x})$ is the trapping potential for atoms in internal state $|j\rangle$, $U_{ij}$ describe the various inter- and intra-state nonlinearities, and $\kappa$ describes the coupling between the two states.

We begin by considering a simplified version of the problem, assuming that only one of these spatial modes is relevant for each of the condensate's components.  This is done by writing the multimode atomic field operators in terms of a set of basis functions $\hat{\psi}_a({\mathbf{r}},t) = \sum_n u_{n}({\mathbf{r}}) \hat{a}_{n}(t)$ and $\hat{\psi}_b({\mathbf{r}},t) = \sum_n u_{n}({\mathbf{r}}) \hat{b}_{n}(t)$, where $\hat{a}_{n}$ and $\hat{b}_{n}$ annihilate a particle with normalised spatial mode $u_n({\mathbf{r}})$ and internal state $|a\rangle$ or $|b\rangle$ respectively. With this expansion, the nonlinear term in the multimode Hamiltonian for the self-interaction of the $|a\rangle$ internal state becomes
\begin{equation}
\hat{H} = \frac{U_{aa}}{2} \int dV \sum_{nmpq} u_n^* u_m^* u_p u_q \, \hat{a}^{\dagger}_{n} \hat{a}^{\dagger}_{m} \hat{a}_{p} \hat{a}_{q} 
\end{equation}
When the system remains predominantly in one mode, we can approximate this sum as the single term $\hat{H} = \hbar\chi_{aa} \hat{a}^{\dagger} \hat{a}^{\dagger} \hat{a} \hat{a}/2$ where
\begin{equation}
\chi_{aa} = \frac{1}{\hbar} U_{aa} \int |u_{i}({\mathbf{r}})|^4 \, dV.
\label{eqChiUequivalence}
\end{equation}
We proceed similarly for the other terms.  In the parameter regimes we are interested in, the detuning between the two spatial modes will be irrelevant, so setting our zero of energy appropriately, we are left with only a standard Kerr-type nonlinearity arising from atom-atom interactions, as well as a linear coupling between the two modes.  The Hamiltonian governing this reduced system is given by
\begin{equation}
\hat{H} = \hbar\frac{\chi_{aa}}{2} \hat{a}^{\dagger} \hat{a}^{\dagger} \hat{a} \hat{a}
          + \hbar\chi_{ab} \hat{a}^{\dagger} \hat{a} \hat{b}^{\dagger} \hat{b}
          + \hbar\frac{\chi_{bb}}{2} \hat{b}^{\dagger} \hat{b}^{\dagger} \hat{b} \hat{b}
          + \hbar\Omega (\hat{a}^{\dagger} \hat{b} + \hat{b}^{\dagger}  \hat{a} )
\label{eqTwoModeHamiltonian}
\end{equation}
where the $\chi_{ij}$ describe the various inter- and intra-mode nonlinearities, and $\Omega$ couples the two modes and allows for population transfer between them.

The system is prepared with all the population in $|a\rangle$, and vacuum in $|b\rangle$. We choose state $|a\rangle$ to be occupied as a coherent state with average number $\langle \hat{a}^{\dagger} \hat{a} \rangle = N$, where $N$ is the total particle number.  As our simulation and experiment are insensitive to the initial phase of the coherent state, this is equivalent to a mixture of coherent states of uncertain phase, or a mixture of Poissonian-distributed number states.  This state is consistent with BEC coherence experiments \cite{Hadzibabic2004}.

At time $t=0$, the coupling $\Omega$ is applied until time $t=t_1$, resulting in a portion of the population being transferred into mode $|b\rangle$. We assume this transfer process is fast relative to the nonlinear energy time scale, ensuring that, as the initial states were coherent states, after the population transfer both $|a\rangle$ and $|b\rangle$ are still described by coherent states with mean number $\langle \hat{a}^{\dagger} \hat{a} \rangle = n_a$ and $\langle \hat{b}^{\dagger} \hat{b} \rangle = n_b$ respectively.  The coupling $\Omega$ is then switched off until time $t_2$, while the atoms interact solely through the nonlinear terms.   After this hold time $\tau_{\mathrm{hold}} = t_2 - t_1 $, the coupling $\Omega$ is switched back on until time $t_3$, with a phase shift $\phi$ compared to its first application. During this last stage, the two modes exchange population and the quadrature variance is converted into number variance, in the same way that homodyne measurements are used in quantum optics to convert quadrature squeezing into number squeezing, which can be directly measured. In this interpretation, $\phi$ is the relative phase of the strong local oscillator, which allows specific phase angles of quadrature squeezing to be examined.  This experiment can also be interpreted as a Ramsey interferometer with a final beam splitter phase of $\phi$.  The entire sequence of pulses and the resulting populations are shown schematically in Figure~\ref{figPulseScheme}.

@@

We might speculate that the effects of these first two problems would be lessened by trap geometries that lead to near-constant density profiles for the BEC.  This should enable higher squeezing.  We cannot make strong conclusions without also considering the effects of coupling between spatial modes, however, which leads to the third potential problem: coupling between modes is inevitable in the presence of a nonlinearity, meaning that the modes cannot be analyzed independently. Coupling between modes will disturb mode-matching, as well as mixing fields with different phase evolution, which can rapidly destroy squeezing.

@@

These stochastic equations take the form
\begin{eqnarray}
\frac{d \phi_a}{dt} = -\frac{i}{\hbar}\left(-\frac{\hbar^2}{2 m}\nabla^2+V_a(\mathbf{x}) + U_{a a} n_a + U_{a b} \tilde{n}_b  \right)\phi_a   + \kappa \phi_b  \nonumber \\
\frac{d \phi_b}{dt} = -\frac{i}{\hbar}\left(-\frac{\hbar^2}{2 m}\nabla^2+V_b(\mathbf{x}) + U_{b b} n_b + U_{a b} \tilde{n}_a   \right)\phi_b   + \kappa^* \phi_a
%\frac{d \phi_a}{dt} = -\frac{i}{\hbar}\left(-\frac{\hbar^2}{2 m}\nabla^2+V_a(\mathbf{x}) + U_{a a} \left( \left|\phi_{a}\right|^2 -\frac{1}{dV}\right) + U_{a b} \left( \left|\phi_b\right|^2 -\frac{1}{2dV}\right)  \right)\phi_a   + \kappa \phi_b  \nonumber \\
%\frac{d \phi_b}{dt} = -\frac{i}{\hbar}\left(-\frac{\hbar^2}{2 m}\nabla^2+V_b(\mathbf{x}) + U_{b b} \left( \left|\phi_{b}\right|^2 - \frac{1}{dV} \right) + U_{a b} \left( \left|\phi_a\right|^2 -\frac{1}{2dV} \right)   \right)\phi_b   + \kappa^* \phi_a
  \label{eqStochasticEquations}
\end{eqnarray}
where $n_j=\left|\phi_{j}\right|^2 -\frac{1}{dV}$, $\tilde{n}_j=\left|\phi_{j}\right|^2 -\frac{1}{2 dV}$, $dV$ is the volume element of the grid on which the simulation is carried out, and the terms proportional to $1/dV$ correspond to vacuum corrections. It is a peculiarity of the Wigner representation that, up to the vacuum correction terms, Eqs.~(\ref{eqStochasticEquations}) look identical to those of the coupled semi-classical nonlinear Schr{\"{o}}dinger equation. The quantum statistical nature of the Wigner equations enters due to the fact that Eqs.~(\ref{eqStochasticEquations}) are run many times and stochastically averaged, with each run using different set of random initial conditions describing (in this case) the noise on a coherent state.

@@

\section{Effects of trap geometries on squeezing in 1D} \label{sec:1Dsim}

@@

\section{Using a $\pi$-pulse to improve mode-matching} \label{sec:pipulse}
Modal mismatch is generated by the different atomic potentials seen by the two internal states.  A straightforward method of minimising this effect is to use a $\pi$-pulse to swap the populations of the internal states after half of the evolution. As can be seen by Eqs~(\ref{eqa1}) and (\ref{eqb1}), such a pulse also leaves the quantum statistics of each state unchanged. In the limit where each state remains in a single mode, and exactly half of the atoms are in each state, this technique ensures that both modes have identical mode shapes after the hold time. In practice this will never work perfectly --- due to the density distributions of $\psi_a(\mathbf{r})$ and $\psi_a(\mathbf{r})$ not reflecting the ground state, resulting in non-identical environments for the two modes, some modal dynamics will occur. Nonetheless, in many regimes, such a $\pi$-pulse can produce some improvement. More importantly, as this scheme accepts that the two modes will change shape, and attempts to make them change in a similar fashion, it allows the possibility of much longer hold times. That is, if we are no longer constrained to hold times short enough such that the mode shapes of the atoms in states $|a \rangle$ and $|b \rangle$ do not change significantly, much more squeezing can be extracted from the system.

These effects are illustrated in Fig.~\ref{piPulseFig}, which shows the best squeezing obtainable from a specific system with and without the $\pi$-pulse, as well as the extra squeezing that can be obtained by allowing a longer hold time as well as the $\pi$-pulse.
\begin{figure}
    \centering
    \includegraphics[width=8cm]{figures/pi_pulse.eps}
    \caption{Effect of a $\pi$-pulse halfway through the hold time. Common parameters used were $n_a = n_b =2 \times 10^5$, $\chi_{aa}=0.03\,\text{s}^{-1}$, $\chi_{ab}=\chi_{bb}=0$. Solid red line: Analytic two-mode solution, $\tau_{\mathrm{hold}}=3\times 10^{-4}$s, $\phi=3.0$ (10.9dB squeezing); Solid blue line: no $\pi$-pulse, $\tau_{\mathrm{hold}}=3\times 10^{-4}$s, $\phi=3.0$ (4.9dB squeezing);  Dashed blue line: with $\pi$-pulse, $\tau_{\mathrm{hold}}=3\times 10^{-4}$s, $\phi=3.05$ (5.8dB squeezing); Dash-dotted blue line: with $\pi$-pulse and the longer hold time $\tau_{\mathrm{hold}}=5\times 10^{-4}$s allowed by the application of the pulse, $\phi=3.87$ (7.2dB squeezing).}
    \label{piPulseFig}
\end{figure}

@@

To minimise the damaging effects of the non-condensed fraction it is better to operate in a regime where the total occupation of the non-zero momentum modes $N_{\mathbf{k}\neq 0}$ is small, which is also fortuitously the regime in which the Bogoliubov theory is a good approximation, so we can investigate in more detail.  In 1D Eq.~\eqref{eqCondensateDepletion} is
\begin{align}
  \frac{N_{\mathbf{k}\neq 0}(t)}{N} &= \frac{(n_a \chi_{aa} \tau_\text{hold} \cos\theta)^2}{N} \sum_{k \neq 0} \sinc^2\left(\sqrt{\omega^0_{k}(\omega^0_{k}+2 \chi_{aa}n_a)}\tau_\text{hold}\right).
\end{align}
We approximate this sum by the integral
\begin{align}
  \frac{N_{\mathbf{k}\neq 0}(t)}{N} &\approx \frac{(n_a \chi_{aa} \tau_\text{hold} \cos\theta)^2}{N} \int_{-\infty}^{+\infty} \sinc^2 \left(\sqrt{\omega^0_{k}(\omega^0_{k}+2 \chi_{aa}n_a)}\tau_\text{hold}\right) \frac{d k}{\Delta k}, \label{eqCondensateDepletionFirstIntegralApproximation}
\end{align}
where $\Delta k = 2\pi/L$ and $L$ is the length of the 1D box condensate.  This approximation will be valid in the limit that $\omega_{k_\text{min}} \tau_\text{hold} \ll 1$, where $k_\text{min} = 2\pi/L$ as for small $k$ the allowed values $k = n_x 2\pi /L$ will closely sample the region $\xi = \sqrt{\omega_k^0(\omega_k^0 + 2 \chi_{aa} n_a)} \tau_\text{hold} < \pi$ where $\sinc^2(\xi)$ is significant.
